---
title: " ADSC 1910 Case Study 1 - Students data by Teachers"
authors:
- name: Shiva Prasad - T00756125
- name: Varsha Tom - T00750156
- name: Keerthi Gopalakrishnan - T00754181
format: pdf
editor: visual
---

## 1 Introduction

### Problem Outline

An elementary school in Kamloops conducted a random survey on grade 5 students to better get to know their students. The elementary school teachers prefer to have a professional second opinion on the quality and structure of the data they have collected.

### Key Methodology Used

The following methods were used to study the data:

1.  **Descriptive Statistics:** To describe the basic features of the data

2.  **Data Visualization:** To visually observe the patterns and trends and highlight outliers

    -   Stem and leaf display for numeric data - To visualize the distribution of Math, Science and English scores

    -   Histogram - To visualize the distribution of given subjects, with density to identify skewness

    -   Bar plots - To visualize the frequency of favourite colours and classes

    -   Box plots - To visualize the distribution of the given subjects and their scores with outlier identification

3.  **Data Cleaning:** To ensure data is consistent

### **Report Structure**

1.  Introduction

2.  Methods Used

3.  Findings and Results

4.  Conclusion

5.  Recommendations

## 2 Methods

1.  **R Programming:** This tool was used to create vectors of the data, clean it and perform descriptive analysis on it

2.  **R Visualization tools:** This tool was used to visualize the data that helped observe the patterns and trends and highlight the outliers in the data

3.  **Quarto:** This tool was used to present the findings with the code and required text in an appealing format

## 3 Findings and Results

The findings and results of this analysis were explained from the basic concepts keeping in mind the audience of the conclusion were elementary school teachers.

The required data was converted into vectors to insert into R.

```{r}
# Creating the required vectors
student_names <- c("Amina", "Hiroshi","Leila","Mateo","Zainab","Rajesh",
                   "Anaya","Omar","Xiang","Sofia","Alejandro","Nia",
                   "Amir","Yara","Kofi")
math_score <- c(88,77,55,63,95,68,75,71,86,59,98,23,85,91,87)
science_score <- c(78,75,68,82,85,71,88,62,76,83,79,68,72,82,77)
english_score <- c(75,64,47,85,98,55,79,65,55,58,89,35,67,95,82)
fav_colour <- c("red","green","blue","NIL","blue","red","blue","NIL"
               ,"green","green","orange","blue","NIL","NIL","red")
fav_class <- c("math","science","gym","gym","english","science",
               "science","math","math","science","english","gym",
               "science","english","math")
```

**Table:**

The table function was used to find the mode in the qualitative data

```{r}
# Using the table function 
table(fav_colour) 
table(fav_class)
```

The mode in fav_colour shows that the observation NIL will have a very high impact on the results of the analysis and it does not contribute to the overall inference from the data.

So, the NIL observations are dropped from the fav_colour vector and the remaining data is stored in a new vector fav_colour_new.

```{r}
fav_colour_new <- fav_colour[fav_colour != "NIL"]
table(fav_colour_new)
```

Now, there are no NIL observations in the fav_colour_new vector and it can be used for analysis.

\newpage

**Stem and Leaf Display:**

Stem and leaf displays of math_score, science_score and english_score were created to display the main features of the distribution and help identify outliers in the data.

```{r}
# Stem and leaf display of quantitative data
stem(math_score,scale=2)
```

The stem and leaf display of math_score showed an outlier (23) in the vector. Most observations of the vector math_score were in the 80s range.

```{r}
stem(science_score,scale=2)
```

The stem and leaf display of science_score showed that most observations were in the 70s range.

\newpage

```{r}
stem(english_score,scale=2)
```

The stem and leaf display of english_score showed that most observations were in the 60s,70s and 80s range.

\newpage

**Histogram:**

Creating the histograms of math_score, science_score and english_score along with their densities to summarize discrete or continuous data measured on an interval scale.

```{r}
# Histogram of quantitative data
hist(math_score,prob=TRUE,xlab = "Math Scores", 
     main = "Histogram of Math scores")
lines(density(math_score),col="red")
```

The histogram of math_score shows a negative skew which shows that most observations in the dataset are in the higher end of the spectrum.

\newpage

```{r}
hist(science_score,prob=TRUE,xlab = "Science Scores", 
     main = "Histogram of Science scores")
lines(density(science_score),col="red")
```

The histogram of science_score shows a slightly negative skew which shows that most observations in the dataset are in the higher end of the spectrum.

\newpage

```{r}
hist(english_score,prob=TRUE,xlab = "English Scores", 
     main = "Histogram of English Scores")
lines(density(english_score),col="red")
```

A bimodal histogram was observed in the english_score vector which shows that there are 2 significant modes in the vector.

\newpage

**Barplot:**

Creating a barplot of fav_colour_new and fav_class to show the relationship between a numeric and a categoric variable.

```{r}
# Barplot of qualitative data
barplot(table(fav_colour_new),xlab = "Favorite colour of students",
        ylab = "Frequency",main = "Barplot of favorite colours of students")
```

Blue is the favourite colour among students with a frequency of 4.

Green and Red are the second favourites at a frequency of 3 and Orange is the colour with the least frequency at 1.

\newpage

```{r}
barplot(table(fav_class),xlab = "Favorite class of students",
        ylab = "Frequency",main = "Barplot of favorite class of students")
```

Science is the favorite class of the students with a frequency of 5 followed by math at a frequency of 4.

English and Gym are next which have the same frequency of 3.

\newpage

**Boxplot:**

Creating the boxplots of math_score, science_score and english_score to show distributions of numeric data values between multiple groups and can also point out if there is any outlier in the data.

```{r}
# Boxplot of quantitative data
boxplot(math_score, ylab = "Math score", 
        main = "Boxplot of Math Score")
```

The boxplot of math_score shows a median of just under 80. There is an outlier in the lower end of the spectrum.

\newpage

```{r}
boxplot(science_score, scale = 2, ylab = "Science score", 
        main = "Boxplot of Science Score")
```

The boxplot of science_score shows a median of around 76.

\newpage

```{r}
boxplot(english_score, scale = 2, ylab = "English score", 
        main = "Boxplot of English Score")
```

The boxplot of english_score shows a median of around 76 and the data is slightly skewed in the positive side.

\newpage

## 4 Discussion / Conclusion

There were missing values in the raw data in the Favourite Colours column. This led to it being low-quality data.

An outlier was found in the Math Scores column which skewed the result.

Math Scores and Science Scores had a negative skew.

English Scores had two significant modes.

Blue was the favourite colour among the students.

Science was the favourite class among the students.

Although some inferences could be made about the students from the data, it cannot be used to represent a large population from this sample. This is mainly because of the absence of some values in the dataset. Another reason is the presence of a few outliers in the data. These two factors hinder us from using this sample data to represent a large population of students.

Nevertheless, since the study’s objective is to learn about the 15 students in the given data, the dataset is pretty good to know about the students.

## 5 Recommendations

While collecting data, make sure not to miss any of the observations as this may hinder the quality of the data.

The quality of the data for analysis decides the relevance of the results from the study of the analyzed data.
